---
title: "Introduction to R"
author: 
   name: "Corrin Fosmire"
   affiliaton: "Rice University"
date: "`r format(Sys.time(), '%B %d   , %Y')`"
runtime: shiny_prerendered
output: learnr::tutorial
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This webpage will run through some exercises to help you get a feel for the power of R. Click on one of the headers on the left to get started!

## Exercise: Arithmetic

Calculate one of the roots of the polynomial \( x^2 + 5*x + 6\). Use the quadratic formula (\(\frac{-b \pm \sqrt{b^2 - 4ac}}{2a}  \)) with a = 1, b = 5, and c = 6.

```{r arithmeticquiz, exercise=TRUE}
#Enter your code here.
```

```{r arithmeticquiz-hint-1}
-5 + sqrt("discriminant")
```

```{r arithmeticquiz-hint-2}
(-5 + sqrt("discriminant")) / 2 * 1
```

```{r arithmeticquiz-hint-3}
(-5 + sqrt(5 ^ 2 - 4 * 1 * 6))
```

```{r arithmeticquiz-hint-4}
(-5 + sqrt(5 ^ 2 - 4 * 1 * 6)) / 2 * 1
```

You should get \( x = -2\) and/or \( x = -3\).

## Exercise: Assignment

Calculate one of the roots of the polynomial \( x^2 + 5*x + 6\). Use the quadratic formula (\(\frac{-b \pm \sqrt{b^2 - 4ac}}{2a}  \)) with a = 1, b = 5, and c = 6. This time, explicitly assign these values to variables called a, b, and c!

```{r assignmentquiz, exercise=TRUE}
#Enter your code here.
```

```{r assignmentquiz-hint-1}
a <- 1
b <- 5
c <- 6
```

```{r assignmentquiz-hint-2}
a <- 1
b <- 5
c <- 6
(-b + sqrt("discriminant")) / 2 * a
```

```{r assignmentquiz-hint-3}
a <- 1
b <- 5
c <- 6
(-b + sqrt(b ^ 2 - 4 * a * c))
```

```{r assignmentquiz-hint-4}
a <- 1
b <- 5
c <- 6
(-b + sqrt(b ^ 2 - 4 * a * c)) / 2 * a
```

You should get \( x = -2\) and/or \( x = -3\).

Now, change the above values of a = 1, b = -6, and c = 9 and rerun the chunk. You should now get \( x = 3\).

## Exercise: Type Conversion

I tried running the chunk below, but I didn't get what I expected.

```{r typefail, echo=TRUE}
sqrt(-1)
```

Convert -1 to a complex number and re-run the chunk to get the complex square root.

```{r typequiz, exercise=TRUE}
#Enter your code here.
```

```{r typequiz-hint-1}
as.complex
```

```{r typequiz-hint-2}
as.complex(-1)
```

```{r typequiz-hint-3}
sqrt(as.complex(-1))
```

You should get the complex square root of \(i\).

## Exercise: Vectors, Lists, Matrices

Make each of the following:

* A vector of your birth year, your birth month, and birth day, all as integers
* A list of your birth year, your birth month, and birth day, all as integers, and your name, as a string. Label all values using names()!
* A matrix of the numbers 1:16, with 4 rows. 
* Change the matrix so the 3rd row is all 3s.

```{r dataquiz, exercise=TRUE}
#Enter your code here.
my_vec <- c()
my_list <- list()
my_mat <- matrix()

list(my_vec, my_list, my_mat)
```

```{r dataquiz-hint-1}
my_vec <- c(1971, 2, 17)
```

```{r dataquiz-hint-2}
my_vec <- c(1971, 2, 17)
my_list <- list(1971, 2, 17, "Jack")
names(my_list) <- c("year", "month", "day", "name")
```

```{r dataquiz-hint-3}
my_vec <- c(1971, 2, 17)
my_list <- list(1971, 2, 17, "Jack")
names(my_list) <- c("year", "month", "day", "name")
my_mat <- matrix(1:16, nrow = 4)
```

```{r dataquiz-hint-4}
my_vec <- c(1971, 2, 17)
my_list <- list(1971, 2, 17, "Jack")
names(my_list) <- c("year", "month", "day", "name")
my_mat <- matrix(1:16, nrow = 4)
my_mat[3,] <- matrix(c(3, 3, 3, 3))
```

## Exercise: Matrix Operations

Use the following formula and matrix multiplication operations to find the 3rd, 4th, and 5th Fibonacci numbers, where \( F_n\) means the \( n\)th Fibonacci number:

\( \begin{pmatrix}
1 & 1 \\
1 & 0
\end{pmatrix}^n  = \begin{pmatrix}
F_{n-1} & F_n \\
F_n & F_{n+1}
\end{pmatrix}\)

```{r matmultquiz, exercise=TRUE}
# Your code here
```

```{r matmultquiz-hint-1}
matrix(c("values"))
```

```{r matmultquiz-hint-2}
matrix(c("values"), nrow=2)
```

```{r matmultquiz-hint-3}
matrix(c(1, 1, 1, 0), nrow=2) %*%
```

```{r matmultquiz-hint-4}
matrix(c(1, 1, 1, 0), nrow=2) %*% matrix(c(1, 1, 1, 0), nrow=2) %*% matrix(c(1, 1, 1, 0), nrow=2) %*% 
matrix(c(1, 1, 1, 0), nrow=2)
```

You should get 2, 3, and 5.

## Exercise: Factors and Data Frames

Create a data frame with three columns: hour, minute, and AM/PM. Enter 5 times. Make sure to use a factor for AM/PM!

```{r dfquiz, exercise=TRUE}
# Your code here
```

```{r dfquiz-hint-1}
data.frame(column=observations)
```

```{r dfquiz-hint-2}
data.frame(hour=c(), minute=c(), ampm=factor())
```

```{r dfquiz-hint-3}
data.frame(hour=c(1), minute=c(5), ampm=factor(c("AM"), levels = c("")))
```

```{r dfquiz-hint-4}
data.frame(hour=c(1, 3), minute=c(5, 30), ampm=factor(c("AM", "PM"), levels = c("AM", "PM")))
```

## Exercise: Reading and Writing Data

Open up RStudio for this exercise. You can't run it in the browser!

We're going to be using the built-in R data set faithful.

First, check the columns of faithful and the first few observations to get a feel for the data.

Next, write the data to faithful.csv with write.csv.

Read it back in with read.csv.

What do you notice about the new data frame? (You can fix this by adding row.names<-FALSE to your write command)

```{r rwquiz, exercise=TRUE}
# Your code here
```

```{r rwquiz-hint-1}
colnames(faithful)
```

```{r rwquiz-hint-2}
colnames(faithful)
head(faithful)
```

```{r rwquiz-hint-3}
colnames(faithful)
head(faithful)

write.csv(faithful, "faithful.csv")
```

```{r rwquiz-hint-4}
colnames(faithful)
head(faithful)

write.csv(faithful, "faithful.csv", row.names=FALSE)
read.csv("faithful.csv")
```