---
title: "Introduction to R"
author: 
   name: "Corrin Fosmire"
   affiliaton: "Rice University"
date: "`r format(Sys.time(), '%B %d, %Y')`"
runtime: shiny_prerendered
output: learnr::tutorial
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This webpage will run through some exercises to help you get a feel for the power of R. Click on one of the headers on the left to get started!

## Exercise: Arithmetic

Calculate one of the roots of the polynomial \( x^2 + 5 \cdot x + 6 \). Use the quadratic formula (\( \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}  \)) with \( a = 1, b = 5, and c = 6 \).

```{r arithmeticquiz, exercise=TRUE}
#Enter your code here.
```

```{r arithmeticquiz-hint-1}
-5 + sqrt("discriminant")
```

```{r arithmeticquiz-hint-2}
(-5 + sqrt("discriminant")) / 2 * 1
```

```{r arithmeticquiz-hint-3}
(-5 + sqrt(5 ^ 2 - 4 * 1 * 6))
```

```{r arithmeticquiz-hint-4}
(-5 + sqrt(5 ^ 2 - 4 * 1 * 6)) / 2 * 1
```

You should get \( x = -2 \) and/or \( x = -3 \).

## Exercise: Assignment

Calculate one of the roots of the polynomial \( x^2 + 5 \cdot x + 6 \). Use the quadratic formula (\(\frac{-b \pm \sqrt{b^2 - 4ac}}{2a}  \)) with \( a = 1, b = 5, and c = 6 \). This time, explicitly assign these values to variables called \( a, b,\) and \( c \)!

```{r assignmentquiz, exercise=TRUE}
#Enter your code here.
```

```{r assignmentquiz-hint-1}
a <- 1
b <- 5
c <- 6
```

```{r assignmentquiz-hint-2}
a <- 1
b <- 5
c <- 6
(-b + sqrt("discriminant")) / 2 * a
```

```{r assignmentquiz-hint-3}
a <- 1
b <- 5
c <- 6
(-b + sqrt(b ^ 2 - 4 * a * c))
```

```{r assignmentquiz-hint-4}
a <- 1
b <- 5
c <- 6
(-b + sqrt(b ^ 2 - 4 * a * c)) / 2 * a
```

You should get \( x = -2 \) and/or \( x = -3 \).

Now, change the above values of \( a = 1, b = -6, \) and \( c = 9 \) and rerun the chunk. You should now get \( x = 3 \).

## Exercise: Type Conversion

I tried running the chunk below, but I was hoping R would give me the closest integer.

```{r roundfail, echo=TRUE}
as.integer(3.9)
```

Use a rounding function to round to the closest integer instead.

```{r typequiz, exercise=TRUE}
#Enter your code here.
```

```{r typequiz-hint-1}
round()
```

```{r typequiz-hint-2}
round(3.9)
```

```{r typequiz-hint-3}
sqrt(as.complex(-1))
```

## Exercise: Vectors and Lists

Make each of the following:

* A vector of your birth year, your birth month, and birth day, all as integers
* A list of your birth year, your birth month, and birth day, all as integers, and your name, as a string. Label all values using ``names()``!

```{r dataquiz, exercise=TRUE}
#Enter your code here.
my_vec <- c()
my_list <- list()

print(my_vec)
print(my_list)
```

```{r dataquiz-hint-1}
my_vec <- c(1971, 2, 17)
```

```{r dataquiz-hint-2}
my_vec <- c(1971, 2, 17)
my_list <- list(1971, 2, 17, "Jack")
```

```{r dataquiz-hint-3}
my_vec <- c(1971, 2, 17)
my_list <- list(1971, 2, 17, "Jack")
names(my_list) <- c("year", "month", "day", "name")
```

## Exercise: Factors and Data Frames

Create a ``data.frame`` with three columns: hour, minute, and AM/PM. Enter 5 times. Make sure to use a ``factor`` for AM/PM!

```{r dfquiz, exercise=TRUE}
# Enter your code here.
```

```{r dfquiz-hint-1}
data.frame(column=observations)
```

```{r dfquiz-hint-2}
data.frame(hour=c(), minute=c(), ampm=factor())
```

```{r dfquiz-hint-3}
data.frame(hour=c(1), minute=c(5), ampm=factor(c("AM"), levels = c("")))
```

```{r dfquiz-hint-4}
data.frame(hour=c(1, 3), minute=c(5, 30), ampm=factor(c("AM", "PM"), levels = c("AM", "PM")))
```

## Exercise: Reading and Writing Data

Open up RStudio for this exercise. You can't run it in the browser!

We're going to be using the built-in R data set ``faithful``.

First, check the columns of ``faithful`` and the first few observations to get a feel for the data.

Next, write the data to ``faithful.csv`` with ``write.csv``.

Read it back in with ``read.csv``.

What do you notice about the new data frame? (You can fix this by adding ``row.names=FALSE`` to your ``write.csv`` command)

```{r rwquiz, exercise=TRUE}
# Enter your code here.
```

```{r rwquiz-hint-1}
colnames(faithful)
```

```{r rwquiz-hint-2}
colnames(faithful)
head(faithful)
```

```{r rwquiz-hint-3}
colnames(faithful)
head(faithful)

write.csv(faithful, "faithful.csv")
```

```{r rwquiz-hint-4}
colnames(faithful)
head(faithful)

write.csv(faithful, "faithful.csv", row.names=FALSE)
read.csv("faithful.csv")
```

## Exercises: Data Manipulation

The ``faithful`` dataset has times between eruptions (the ``waiting`` column) measured in minutes. Add a new column giving the waiting time in hours.

```{r newcolquiz, exercise=TRUE}
# Enter your code here.
```

```{r newcolquiz-hint-1}
faithful$waiting / 60
```

```{r newcolquiz-hint-2}
faithful$waiting.hrs <- faithful$waiting / 60
```

Find the times when the waiting time is more than 30x as long as the eruption. There should be 17, starting with 17 and ending with 263. Save it into the variable ``longwaits``.

```{r subsetquiz, exercise=TRUE}
longwaits <- c()

print(longwaits)
print(nrow(longwaits))
```

```{r subsetquiz-hint-1}
longwaits <- subset(faithful, "condition")
```

```{r subsetquiz-hint-2}
longwaits <- subset(faithful, waiting > 30 * eruptions)
```

I'm giving you some data that has missing values. Compute the mean and standard deviation of the values that are still there. Don't manually change the values in ``data`` - this approach won't be convenient when you deal with larger data sets! You should get 11.66667 and 15.14376.

```{r naquiz, exercise=TRUE}
data <- c(1, 5, NA, 29, NA)

# Enter your code here.
```

```{r naquiz-hint-1}
c(mean(data), sd(data))
"What is wrong with this?"
```

```{r naquiz-hint-2}
c(mean(data, na.rm=TRUE), sd(data, na.rm=TRUE))
```

We're going to be exploring the ``iris`` data set, which contains observations of three different species of irises, as well as petal length and width and sepal length and width. 

If you're interested in what all these words mean or what the flowers in question look like, check out http://www.lac.inpe.br/~rafael.santos/Docs/R/CAP394/WholeStory-Iris.html !

Use ``tapply`` to find the median petal length (``Petal.Length``) of each species (``Species``)! You should find setosa's median is 1.50, versicolor's is 4.35, and virginica's is 5.55.

Food for thought: do you think you can tell which species an observation belongs to just based on the petal and sepal measurements? This is a common task in machine learning!

```{r tapplyquiz, exercise=TRUE}
# Enter your code here.
```

```{r tapplyquiz-hint-1}
tapply("", "", median)
```

```{r tapplyquiz-hint-2}
tapply(iris$Petal.Length, "", median)
```

```{r tapplyquiz-hint-3}
tapply(iris$Petal.Length, iris$Species, median)
```

## Exercises: Plotting & Modeling

Make a boxplot and a histogram of Old Faithful's ``waiting`` time, using Scott's breaks!

```{r boxhistquiz, exercise=TRUE}
# Enter your code here.
```

```{r boxhistquiz-hint-1}
boxplot(faithful$waiting)
```

```{r boxhistquiz-hint-2}
boxplot(faithful$waiting)
hist(faithful$waiting)
```

```{r boxhistquiz-hint-3}
boxplot(faithful$waiting)
hist(faithful$waiting, breaks="scott")
```

Now, make a scatterplot of the faithful data set, with the x-axis as eruptions (the duration of an eruption) and the y-axis as waiting (time to next eruption).

```{r scatterquiz, exercise=TRUE}
# Enter your code here.
```

```{r scatterquiz-hint-1}
plot("x-axis", "y-axis")
```

```{r scatterquiz-hint-2}
plot(faithful$eruptions, faithful$waiting) # or just plot(faithful)
```

Make a linear model (use the ``lm`` function) predicting waiting time from eruption length and plot it on the graph. Does it look like a good fit?

```{r modelquiz, exercise=TRUE}
# Copy your answer from last exercise here!

# Add the line of best fit!
```

```{r modelquiz-hint-1}
plot(faithful)
lm(faithful$waiting ~ faithful$eruptions)
```

```{r modelquiz-hint-2}
plot(faithful)
lm(faithful$waiting ~ faithful$eruptions)
abline()
```

```{r modelquiz-hint-3}
plot(faithful)
model <- lm(faithful$waiting ~ faithful$eruptions)
abline(model)
```

Congratulations! You've passed the Introduction to R.

## Reference

It's a good idea to get in the habit of using RStudio's amazing cheatsheets. Check out [this one](https://github.com/rstudio/cheatsheets/blob/master/base-r.pdf) for most of what we talked about today.