---
title: "R Visualization and Data Manipulation"
author: 
   name: "Corrin Fosmire"
   affiliaton: "Rice University"
date: "`r format(Sys.time(), '%B %d, %Y')`"
runtime: shiny_prerendered
output: learnr::tutorial
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(viridis)
library(datasets)

knitr::opts_knit$set(root.dir = "/home/infrared/datademoes/")
knitr::opts_chunk$set(root.dir = "/home/infrared/datademoes/")
tidystates <- read_csv("/home/infrared/datademoes/tidystates.csv")
```

## Introduction

This webpage will run through some exercises to help get you comfortable using the most modern techniques in R!

## Exercises: tibbles

Convert the data frame ``state.x77`` to a ``tibble``. Then, convert it back.

```{r tibbleaquiz, exercise=TRUE}
state.x77
```

```{r tibbleaquiz-hint-1}
state77tibble <- as_tibble(state.x77)
```

```{r tibbleaquiz-hint-2}
state77tibble <- as_tibble(state.x77)
state77df <- as.data.frame(state77tibble)
```

Make the following data using a ``tribble``:

<table width=400>
<tr>
 <th>Month</th>
 <th>Savings</th>
</tr>
<tr>
 <td>January</td>
 <td>$100</td>
</tr>
<tr>
 <td>February</td>
 <td>$150</td>
</tr>
</table> 

```{r tribbleaquiz, exercise=TRUE}
#Enter your code here.
```

```{r tribbleaquiz-hint-1}
tribble(
  ~column,
  value
)
```

```{r tribbleaquiz-hint-2}
tribble(
  ~Month, ~Savings,
  "January", 100,
  "February", 150
)
```
  
## Exercises: tidying data

Here's some data that might or might not be real. Regardless, it's not tidy, so take a look at it, and tidy it up!

Here, try ``spread``ing the data out:

```{r spreadquiz-setup, include=FALSE}
very_real_data <- tibble(Department=c("Computer Science", "Computer Science", "Statistics", "Statistics", "English", "English"), type=c("Enrollment", "Faculty", "Enrollment", "Faculty", "Enrollment", "Faculty"), value=c(500, 20, 400, 15, 300, 10))
```

```{r spreadquiz, exercise=TRUE}
very_real_data
```

```{r spreadquiz-hint-1}
spread(very_real_data, "key column", "value column")
```

```{r spreadquiz-hint-2}
spread(very_real_data, type, value)
```

Now, try to ``gather`` the data to make it tidy! Note that since the data has numerals (years) as column names, you will have to reference it using backticks, as in `2019`.

```{r gatherquiz-setup, include=FALSE}
real_data <- tibble(Department=c("Computer Science", "Statistics", "English"), `2018`=c(300, 400, 500), `2019`=c(500, 400, 300))
```

```{r gatherquiz, exercise=TRUE}
real_data
```

```{r gatherquiz-hint-1}
gather(real_data, "columns", key="?", value="?")
```

```{r gatherquiz-hint-2}
gather(real_data, `2018`, `2019`, key="?", value="?")
```

```{r gatherquiz-hint-3}
gather(real_data, `2018`, `2019`, key="Year", value="Enrollment")
```

## Exercises: readr

First, read in the state data csv from ``tidystates.csv``. Make sure to use the readr ``read_csv`` and not the base R ``read.csv`` to get a ``tibble`` immediately! Then, import the fixed width file ``ufo.fwf`` with the widths of 6, 50, 15, 4, 4, 4, and 4. Label the columns as Year, Name, Location, PhysicalEffects, Multimedia, ETContact, and Abduction.

```{r importquiz, exercise=TRUE}
#Enter your code here.
```

```{r importquiz-hint-1}
read_csv
read_fwf

ufofwf
```

```{r importquiz-hint-2}
read_csv("tidystates.csv")
ufofwf <- read_fwf("ufo.fwf", fwf_widths(c(6, 50, 15, 4, 4, 4, 4)))

ufofwf
```

```{r importquiz-hint-3}
read_csv("tidystates.csv")
ufofwf <- read_fwf("ufo.fwf", fwf_widths(c(6, 50, 15, 4, 4, 4, 4)))
colnames(ufofwf) <- c("Year", "Name", "Location", "PhysicalEffects", "Multimedia", "ETContact", "Abduction")

ufofwf
```

## Exercises: purrr

Use ``map2`` to format the following message for each row of the ``tidystates`` tibble: STATE has population POPULATION! Replace the two arguments in the printing command.

```{r map2quiz, exercise=TRUE}
# Use the printing command:
#    sprintf("%s has population %d!", STATE, POPULATION)
# but replace STATE and POPULATION!

```

```{r map2quiz-hint-1}
map2("column1", "column2", ~ sprintf("%s has population %d!", .x, .y))
```

```{r map2quiz-hint-2}
map2(tidystates$State, tidystates$Population, ~ sprintf("%s has population %d!", .x, .y))
```

Use ``keep`` to take only the state populations with populations above 2 million (note that population is here measured in 1000s, so this is 2000). Use ``discard`` to remove the state areas under 50000 square miles.

```{r keepdiscardquiz, exercise=TRUE}
# Enter your code here.
```

```{r keepdiscardquiz-hint-1}
purrr::keep(tidystates$Population, "condition formula")
discard(tidystates$Area, "condition formula")
```

```{r keepdiscardquiz-hint-2}
purrr::keep(tidystates$Population, ~ .x > 2000)
discard(tidystates$Area, ~ .x < 50000)
```

Check to see whether all the illiteracy rate of every state is under 2.5% (use ``every``). Check to see whether there is a state with a murder rate of higher than 15 per 100,000 (use ``some``).

```{r everysomequiz, exercise=TRUE}
# Enter your code here.
```

```{r everysomequiz-hint-1}
every(tidystates$Illiteracy, "condition formula")
some(tidystates$Murder, "condition formula")
```

```{r everysomequiz-hint-2}
every(tidystates$Illiteracy, ~.x < 2.5)
some(tidystates$Murder, ~.x > 15)
```

## Exercise: dplyr

We're going to be using our ``tidystates`` data to practice our new operations. 

Filtering: Get me all the states which have illiteracy rates less than or equal to 1%.

```{r filterquiz, exercise=TRUE}
# Enter your code here.
```

```{r filterquiz-hint-1}
filter(tidystates, "condition")
```

```{r filterquiz-hint-2}
filter(tidystates, Illiteracy <= 1)
```

Mutating: Add a new column giving the population density by taking ``Population`` divided by ``Area``. Use ``mutate`` first, then change it to ``transmute`` to observe the difference.

```{r mutatequiz-setup}
tidystates <- filter(tidystates, Illiteracy <= 1)
```

```{r mutatequiz, exercise=TRUE}
# Enter your code here.
```

```{r mutatequiz-hint-1}
mutate(tidystates, "operation")
```

```{r mutatequiz-hint-2}
mutate(tidystates, Density = Population / Area)
```

Arrange: Sort the states from highest to lowest murder rate.

```{r arrangequiz, exercise=TRUE}
# Enter your code here.
```

```{r arrangequiz-hint-1}
arrange(tidystates, "column")
```

```{r arrangequiz-hint-2}
arrange(tidystates, desc(Murder))
```

Join: Add in the region using the ``regions`` data frame I'm giving you. Do an ``inner_join`` on ``State``!

```{r joinquiz-setup}
tidystates <- filter(tidystates, Illiteracy <= 1)
tidystates <- mutate(tidystates, Density = Population / Area)
regions <- tibble(State=state.name, Region=state.region)
```

```{r joinquiz, exercise=TRUE}
# Enter your code here.
```

```{r joinquiz-hint-1}
inner_join(tidystates, regions, by="?")
```

```{r joinquiz-hint-2}
inner_join(tidystates, regions, by="State")
```

Group By: Group states by their region and count how many are in each region.

```{r groupbyquiz-setup}
tidystates <- filter(tidystates, Illiteracy <= 1)
tidystates <- mutate(tidystates, Density = Population / Area)
regions <- tibble(State=state.name, Region=state.region)
tidystates <- inner_join(tidystates, regions, by="State")
```

```{r groupbyquiz, exercise=TRUE}
# Enter your code here.
```

```{r groupbyquiz-hint-1}
group_by(tidystates, Region)
```

```{r groupbyquiz-hint-2}
group_by(tidystates, Region) %>%
  count()
```

Summarize: Now, calculate the average population density of each region.

```{r summarizequiz-setup}
tidystates <- filter(tidystates, Illiteracy <= 1)
tidystates <- mutate(tidystates, Density = Population / Area)
regions <- tibble(State=state.name, Region=state.region)
tidystates <- inner_join(tidystates, regions, by="State")
tidystates <- group_by(tidystates, Region)
```

```{r summarizequiz, exercise=TRUE}
# Enter your code here.
```

```{r summarizequiz-hint-1}
summarize(tidystates, "operation")
```

```{r summarizequiz-hint-2}
summarize(tidystates, avgdensity = mean(Density))
```

Now, do all that in one go with the pipe operator! (%>%) You don't need to do the arrange, since it won't change the answer.

```{r pipequiz-setup}
regions <- tibble(State=state.name, Region=state.region)
```

```{r pipequiz, exercise=TRUE}
tidystates %>%
```

```{r pipequiz-solution}
tidystates %>%
  filter(Illiteracy <= 1) %>%
  mutate(Density = Population / Area) %>%
  inner_join(regions, by="State") %>%
  group_by(Region) %>%
  summarize(avgdensity = mean(Density))
```

## Exercises: ggplot

```{r ggplotsetupchunk, include=FALSE, echo=FALSE}
regions <- tibble(State=state.name, Region=state.region)
tidystates <- tidystates %>%
  mutate(Density = Population / Area) %>%
  inner_join(regions, by="State")
```

Make a scatterplot of state ``Population`` by ``Area``.

```{r ggscatterquiz, exercise.setup="ggplotsetupchunk", exercise=TRUE}
ggplot(data = tidystates) +
```

```{r ggscatterquiz-hint-1}
ggplot(data = tidystates) +
  geom_point(mapping = "?")
```

```{r ggscatterquiz-hint-2}
ggplot(data = tidystates) +
  geom_point(mapping = aes(x = Area, y = Population))
```

Improve that plot by adding a color indicating which ``Region`` each state is in!

```{r ggscatter2quiz, exercise.setup="ggplotsetupchunk", exercise=TRUE}
ggplot(data = tidystates) +
```

```{r ggscatter2quiz-hint-1}
ggplot(data = tidystates) +
  geom_point(mapping = "?")
```

```{r ggscatter2quiz-hint-2}
ggplot(data = tidystates) +
  geom_point(mapping = aes(x = Area, y = Population, color = Region))
```

Make a histogram and a boxplot of state ``Income``. Note that histogram maps ``x``, while boxplot maps ``y``!

```{r gghistboxquiz, exercise.setup="ggplotsetupchunk", exercise=TRUE}
ggplot(data = tidystates) +
```

```{r gghistboxquiz-hint-1}
ggplot(data = tidystates) +
  geom_histogram(mapping = "?")

ggplot(data = tidystates) +
  geom_boxplot(mapping = "?")
```

```{r gghistboxquiz-hint-2}
ggplot(data = tidystates) +
  geom_histogram(mapping = aes(x = Income))

ggplot(data = tidystates) +
  geom_boxplot(mapping = aes(y = Income))
```

Use facets to visualize our earlier scatter plot by region. Note how this shoes the data of our colored scatterplot in a slightly different way!

```{r ggscatter3quiz, exercise.setup="ggplotsetupchunk", exercise=TRUE}
ggplot(data = tidystates) +
```

```{r ggscatter3quiz-hint-1}
ggplot(data = tidystates) +
  geom_point(mapping = "?")
```

```{r ggscatter3quiz-hint-2}
ggplot(data = tidystates) +
  geom_point(mapping = aes(x = Area, y = Population)) +
  facet_wrap(~ Region, nrow = 2)
```

Add a main title to our colored scatterplot!

```{r ggscatter4quiz, exercise.setup="ggplotsetupchunk", exercise=TRUE}
ggplot(data = tidystates) +
```

```{r ggscatter4quiz-hint-1}
ggplot(data = tidystates) +
  geom_point(mapping = aes(x = Area, y = Population, color = Region)) +
  labs()
```

```{r ggscatter4quiz-hint-2}
ggplot(data = tidystates) +
  geom_point(mapping = aes(x = Area, y = Population, color = Region)) +
  labs(title = "Area and Population of U.S. States")
```

Make our scatterplot again, this time using ``Income`` as our color mapping with ``scale_color_viridis``!

```{r ggscatter5quiz, exercise.setup="ggplotsetupchunk", exercise=TRUE}
ggplot(data = tidystates) +
```

```{r ggscatter5quiz-hint-1}
ggplot(data = tidystates) +
  geom_point(mapping = aes(x = Area, y = Population, color = "?")) +
  scale_color_viridis() +
  labs(title = "Area and Population of U.S. States")
```

```{r ggscatter5quiz-hint-2}
ggplot(data = tidystates) +
  geom_point(mapping = aes(x = Area, y = Population, color = Income)) +
  scale_color_viridis() +
  labs(title = "Area and Population of U.S. States")
```

Now, try out the themes for our plot. They are

* theme_bw()
* theme_classic()
* theme_dark()
* theme_gray()
* theme_light()
* theme_linedraw()
* theme_minimal()
* theme_void()

Just pick one you think looks nice!

```{r ggscatter6quiz, exercise.setup="ggplotsetupchunk", exercise=TRUE}
ggplot(data = tidystates) +
```

```{r ggscatter6quiz-solution}
ggplot(data = tidystates) +
  geom_point(mapping = aes(x = Area, y = Population, color = Income)) +
  scale_color_viridis() +
  labs(title = "Area and Population of U.S. States") +
  theme_bw() #or anything else!
```

## Exercises: everything before

For your Final Exam, I'm giving you a new data set, called ``agdata``. It measures the total agricultural production by county. Do all of the following, using as much %>% and + as you can:

* Group the data by state.
* Sum all the value produced by each state.
* Perform an inner join with our ``tidystates`` data on the ``State`` column.
* Send the data to ggplot.
* Add a scatterplot layer.
* Add a title to the plot.
* Use the dark theme.

Each hint will give you one step.

```{r finalexam, exercise=TRUE}
# Enter your code here.
```

```{r finalexam-hint-1}
agdata %>%
  group_by(State)
```

```{r finalexam-hint-2}
agdata %>%
  group_by(State) %>%
  summarize(total = sum(Value))
```

```{r finalexam-hint-3}
agdata %>%
  group_by(State) %>%
  summarize(total = sum(Value)) %>%
  inner_join(tidystates, by="State") 
```

```{r finalexam-hint-4}
agdata %>%
  group_by(State) %>%
  summarize(total = sum(Value)) %>%
  inner_join(tidystates, by="State") %>% 
  ggplot()
```

```{r finalexam-hint-5}
agdata %>%
  group_by(State) %>%
  summarize(total = sum(Value)) %>%
  inner_join(tidystates, by="State") %>% 
  ggplot() +
  geom_point(size=1, shape=23, mapping = aes(Income, total)) 
```

```{r finalexam-hint-6}
agdata %>%
  group_by(State) %>%
  summarize(total = sum(Value)) %>%
  inner_join(tidystates, by="State") %>% 
  ggplot() +
  geom_point(size=1, shape=23, mapping = aes(Income, total)) +
  labs(x = "Income of State", y = "Total Agricultural Production of State",
       title = "Income and Total Agricultural Production by State")
```


```{r finalexam-hint-7}
agdata %>%
  group_by(State) %>%
  summarize(total = sum(Value)) %>%
  inner_join(tidystates, by="State") %>% 
  ggplot() +
  geom_point(size=1, shape=23, mapping = aes(Income, total)) +
  labs(x = "Income of State", y = "Total Agricultural Production of State",
       title = "Income and Total Agricultural Production by State") +
  theme_bw()
```

Congratulations! You're now an expert at data manipulation and visualization in R.

## Reference

It's a good idea to get in the habit of using RStudio's amazing cheatsheets. You can find them in the Help -> Cheatsheets menu. Here are a few extremely useful ones, for today and beyond:

* [readr & tidying (importing)](https://github.com/rstudio/cheatsheets/blob/master/data-import.pdf)
* [dplyr (data manipulation)](https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf)
* [purrr (vector manipulation)](https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf)
* [ggplot2 (plotting)](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf)